---
title: "Impact of Mother's Characteristics on Child Test Scores"
date: 2026-02-15
description: |
  Exploring the NLSY dataset and observing how maternal IQ and family advantage 
  work together to influence child development. 
output:
  distill::distill_article:
    self_contained: false
    code_folding: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
kidiq <- read.csv("kidiq.csv")
```
## Overview

In this project, I wanted to see how much a child’s home environment affects 
their success. I examined how maternal education and employment status, grouped as ‘family advantage’ interact with maternal IQ to shape child test scores as an outcome of intergenerational inequality.

## The Double Advantage
The graph below breaks down these test scores. I grouped the families by their
advantage level and separated them by the mother's IQ to see if having a more
"advantaged" background makes a difference even when IQ is the same.

```{r plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6, out.width="100%"}

plot_iq <- kidiq %>%
  mutate(
    family_advantage = case_when(
      mom_hs == 0 & mom_work <= 2 ~ "Low advantage",
      mom_hs == 0 & mom_work >  2 ~ "Mixed advantage",
      mom_hs == 1 & mom_work <= 2 ~ "Mixed advantage",
      mom_hs == 1 & mom_work >  2 ~ "High advantage",
      TRUE ~ NA_character_
    ),
    iq_group = case_when(
      mom_iq <= quantile(mom_iq, 0.33, na.rm = TRUE) ~ "Lower IQ Group",
      mom_iq <= quantile(mom_iq, 0.67, na.rm = TRUE) ~ "Middle IQ Group",
      TRUE ~ "Higher IQ Group"
    ),
    family_advantage = factor(family_advantage,
                              levels = c("Low advantage","Mixed advantage","High advantage")),
    iq_group = factor(iq_group,
                      levels = c("Lower IQ Group","Middle IQ Group","Higher IQ Group"))
  ) %>%
  filter(!is.na(family_advantage), !is.na(iq_group)) %>%
  
  ggplot(aes(x = family_advantage, y = kid_score, fill = family_advantage)) +
  geom_violin(trim = FALSE, alpha = 0.6, width = 0.8, color = NA) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black", fatten = 1) +
  geom_jitter(width = 0.10, height = 0, alpha = 0.25, size = 1.2, color = "black", show.legend = FALSE) +
  facet_wrap(~ iq_group, nrow = 1) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_x_discrete(labels = c("Low advantage"="Low","Mixed advantage"="Mixed","High advantage"="High")) +
  labs(
    title = "Child Scores by Family Background and Maternal IQ",
    subtitle = "Higher family advantage leads to higher scores even within the same IQ groups.",
    x = "Family Advantage Level",
    y = "Child's Test Score",
    caption = "Data: NLSY (Gelman & Hill, 2007)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16),
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.major.x = element_blank()
  )

print(plot_iq)
```

## Understanding the Results 

The plot suggests that child test performance is not explained by maternal IQ alone, but is strongly conditioned by the family’s socioeconomic position. In the lower and middle IQ groups, there is a clear upward trend: as family advantage increases, child test scores rise accordingly. This indicates that institutional and socioeconomic resources associated with maternal education and employment status function as key mediating factors that can significantly improve developmental outcomes.

A key finding in this visualisation is the equalising effect of a high-advantage home. Children born to mothers in the lower IQ group who have high family advantage often achieve scores comparable to children in the middle IQ group. This suggests that differences in child outcomes are shaped not only by maternal cognitive ability, but also by the distribution of socioeconomic advantage across households, reflecting broader patterns of inequality reproduction within the education system. Whilst the impact of family advantage appears to level off for mothers who already have high IQs, the overall pattern indicates that stratification in outcomes is partially conditioned by structural socioeconomic positioning, particularly for those not starting from a high-IQ baseline.

## The Myth of Meritocracy?

While the violin plots show general group trends, this section explores whether "merit" (maternal IQ) alone determines outcomes, or if structural factors like education create the true path to success.

```{r plot2, echo=FALSE, message=FALSE, warning=FALSE}
plot_merit <- kidiq %>%
  mutate(
    mom_hs_label = factor(mom_hs, labels = c("No HS diploma", "HS diploma")),
    ability_level = case_when(
      mom_iq <= quantile(mom_iq, 0.25, na.rm = TRUE) ~ "Low ability",
      mom_iq <= quantile(mom_iq, 0.75, na.rm = TRUE) ~ "Medium ability",
      TRUE ~ "High ability"
    ),
    outcome_level = case_when(
      kid_score <= quantile(kid_score, 0.25, na.rm = TRUE) ~ "Low outcome",
      kid_score <= quantile(kid_score, 0.75, na.rm = TRUE) ~ "Medium outcome",
      TRUE ~ "High outcome"
    ),
    surprising_case = case_when(
      ability_level == "High ability" & outcome_level == "Low outcome" ~ 
        "High ability, low outcome",
      ability_level == "Low ability"  & outcome_level == "High outcome" ~
        "Low ability, high outcome",
      TRUE ~ NA_character_
    )
  )

ggplot(plot_merit, aes(mom_iq, kid_score)) +
  stat_bin2d(bins = 24) +
  scale_fill_viridis_c(name = "Cases", option = "C", direction = -1) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed",
              color = "white", linewidth = 1) +
  geom_point(
    data = dplyr::filter(plot_merit, !is.na(surprising_case)),
    aes(color = surprising_case),
    size = 2.6
  ) +
  scale_color_manual(
    values = c("High ability, low outcome" = "#E69F00",
               "Low ability, high outcome" = "#56B4E9"),
    name = "Surprising cases"
  ) +
  facet_wrap(~ mom_hs_label, nrow = 1) +
  coord_cartesian(xlim = c(75, 135), ylim = c(40, 120)) +
  labs(
    title = "Is Merit Enough?",
    subtitle = "Tiles show where most cases sit; dashed lines show the IQ→score trend. 
    Highlighted dots mark exceptions where merit alone didn’t predict outcomes.",
    x = "Mother’s IQ (ability proxy)",
    y = "Child’s test score",
    caption = "Data: NLSY (Gelman & Hill, 2007)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.margin = margin(10, 20, 10, 20)
  )
```
Whilst the initial analysis suggests that background advantage is associated with higher scores, a closer examination reveals that this relationship is institutionally mediated. Within the high-school completion group, the steeper trend indicates that maternal ability translates into child outcomes more strongly under conditions of completed educational attainment.

The most revealing feature of the visualisation is the presence of outlier cases. Several children achieve high scores despite lower maternal ability, and these cases occur more frequently when mothers hold a high school diploma, a credential produced through public education systems. By contrast, higher maternal ability does not consistently yield strong outcomes in the absence of this educational threshold. These patterns suggest that individual ability operates within structurally conditioned pathways, with educational attainment shaping the distribution of early-life outcomes through formalised systems of stratification.

[← Back to Projects](../../projects.html)

